---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
<<<<<<< HEAD
editor_options: 
  chunk_output_type: console
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
<<<<<<< HEAD
mtcars
cars
###mtcars is another one
=======
cars
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed
<<<<<<< HEAD
mtcars$cyl
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
<<<<<<< HEAD

is.vector(car_speed)
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed)
<<<<<<< HEAD
#### converting the vector into a data frame
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist
<<<<<<< HEAD
#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
#only dataframes will be created
=======

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
<<<<<<< HEAD
######cbind: Take a sequence of vector, matrix or data-frame arguments and combine by columns or rows, respectively.
?cbind
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
<<<<<<< HEAD
#data.frame is for creating a new dataframe
=======
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  
nrow(df)  
length(df)  #same as ncol
<<<<<<< HEAD
summary(df) # great for seeing whats going on
head(df) #show the first 6 rows of df
head(df,50)

#If you know the number of the column you want you can refer to that to access column
df[,1]
#matrix notation: [,1] 
##[rows, columns], if nothing, gives everything
df[1,]
df[1:10,2] #slice
df[c(1,2,3),] ###?? what does this do?
## getting everything from row 1 2 3. c is to combine row 1 2 3
df2 <- df[c(2,1,3),]$car_speed
#you could also use this notation to delete columns

=======
summary(df)
head(df) #show the first 6 rows of df

#If you know the number of the column you want you can refer to that to access column
df[,1]

#you could also use this notation to delete columns
>>>>>>> 5e9b12925d958201c855d73439002d288857353b
df <- df[,-2]
df
```
