---
title: "Assignment 2: Coding Basics"
author: "Queenie Wei"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 

seqonetothirtybythree <- seq(1,30,3)
seqonetothirtybythree
# this is to generate a sequence that goes from 1 to 30, increasing by 3s.

#2. 

mean(seqonetothirtybythree)
median(seqonetothirtybythree)

### This is for calculating the mean of the sequence in question 1

#3. 

mean(seq(1,30,3)) > median(seq(1,30,3))


#### The code above tries to see if the mean of the sequence in question 1 is larger than the median of the sequence.


```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
### 5
## 5a
names <- as.vector(c("fudge", "muffin", "seal", "kangaroo")) ###vector type: char
names                            ##typeof(names)
## 5b
test_score <- as.vector(c(100, 23, 76, 10))  
test_score <- as.integer(test_score)
test_score                            ##typeof(testscore)
####### vector type: originally: double, after conversion: integer
## 5c
passed <- as.vector((c(TRUE, FALSE, TRUE, FALSE))) ####vector type:logical
passed                            ##typeof(passed)  
## alternatively
x <- c()
for (i in test_score){
  x <- append(x, i>50)
}
##x    ####vector type:logical
### 7
studentpassfail <-data.frame(names, test_score, passed)
colnames(studentpassfail) <- c("Names", "Test_scores", "Passed")
studentpassfail
#### this code creates a data frame called studentpassfail and combines the three vectors from above into a dataframe

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Dataframes can contain different classes of data, whereas matrices can only have one kind of data. since we have different classes of data, a matrix would not be suitable in this situation.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

didtheypass <- function(scores){
  for (i in scores){
   if (i >=  50)
    print("TRUE")
   else
    print("FALSE")
  }
}
didtheypass(test_score)
didtheypass2 <- function(scores){
  result <- ifelse(scores >= 50, "TRUE", "FALSE")
  print(result)
}
didtheypass2(test_score)
#####
# works cited:
#   chatgpt
# prompt:
#   i fed chatgpt my code for if and else, and used the prompt "rewrite it with ifelse() so that when scores are equal or over fifty, "TRUE" is printed, otherwise, "FALSE" is printed"
# ####didtheypass2
####
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Both worked, with ifelse being more efficient. However, ifelse RETURNS results, and the way chatgpt did it is a very sneaky way and doesnt 100% do the job.
